---
title: "The Association Between Colleges’ Median Earnings with Instructional Expenses and Control Type of Colleges"
author: "Avianna Bui"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1. Load Packages & Data Set

## Load Packages
```{r, message = FALSE}
library(readr) 
library(dplyr) 
library(ggplot2) 
library(ggmosaic) 
library(broom) 
library(forcats) 
```

## Load Data

```{r, message = FALSE}
collegescorecard <- read_csv("Scorecard.csv") 
```

# Part 2. Cleaning & Data Context

## Data Cleaning

```{r}
scorecard <- collegescorecard %>%
    filter(INEXPFTE > 0) %>% 
    filter(!is.na(MD_EARN_WNE_P6)) %>% 
    mutate(PREDDEG.Cat = case_when(
    PREDDEG==1 ~ "Certificate degrees", 
    PREDDEG==2 ~ "Associate degrees", 
    PREDDEG==3 ~ "Bachelor's degrees")) %>%  
    mutate(PREDDEG.Cat = fct_relevel(PREDDEG.Cat, "Certificate degrees")) %>% 
    mutate(centeredexp = (INEXPFTE) - mean(INEXPFTE)) %>% 
    mutate(centeredfaminc = (FAMINC) - mean(FAMINC)) %>%
    mutate(aboveaveincome = ifelse(MD_EARN_WNE_P6 > 49785, 1, 0)) %>% 
    mutate(Control.Cat = case_when(
    CONTROL == 1 ~ "Public", 
    CONTROL == 2 | CONTROL == 3 ~ "Private")) %>%  
    mutate(Control.Cat = fct_relevel(Control.Cat, "Public")) %>% 
    mutate(FAMINC1000 = FAMINC/1000) %>%
    mutate(FAMINC.Cat = cut(FAMINC1000, 3))
```

## Data Context

```{r}
scorecard %>%
    summarize(min(INEXPFTE), max(INEXPFTE),
            min(MD_EARN_WNE_P6), max(MD_EARN_WNE_P6),
            min(FAMINC), max(FAMINC))
```

# Part 3. Research Question 1 - Do Higher College Instructional Expenditures per Full-time Student Correspond with Higher Earnings After Graduation?

## Visualization

```{r, message = FALSE}
scorecard %>%
    ggplot(aes(x = centeredexp, y = MD_EARN_WNE_P6, color = PREDDEG.Cat)) +
    geom_point(alpha = 0.4) + 
    geom_smooth(method = 'lm', se = FALSE) + 
    labs(title = "Median Earnings vs. Centered Instructional Expenses", x = "Centered Instructional Expenses per Full-time Student (USD)", y = "Median Earnings 6 Years After Entry (USD)", color = "Predominant Degree Awarded") + 
    scale_color_brewer(palette = "Dark2") +
    theme_classic() + 
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

# calculate correlation
scorecard %>%
    group_by(PREDDEG.Cat) %>%
    summarize(cor(centeredexp,MD_EARN_WNE_P6)) 

# calculate instructional expenditures range
scorecard %>%
    group_by(PREDDEG.Cat) %>%
    summarize(min(INEXPFTE), max(INEXPFTE)) 
```

Overall, we can see from the graph that instructional expenses have a weakly to moderately positive association with colleges’ median earnings after graduation: institutions with higher instructional expenditures per full-time student are estimated to have a higher median income among graduates. Specifically, among institutions with 3 different primary degree types, the greatest correlation between instructional expenses per student and median earnings occurs in predominantly bachelor’s degree colleges with a 0.45 correlation coefficient, indicating a moderately, positive relationship. The correlation coefficients for institutions with primarily associate and certificate degrees are respectively 0.29 and 0.26, suggesting a weak to moderate positive correlation. 

Colleges that predominantly offer bachelor’s degrees appear to have higher median earnings 6 years after college enrollment. From the data points in the visualization, we can approximate that the majority of median earnings for predominantly bachelor’s degree institutions ranges from $30,000 to $75,000, compared to institutions with primarily associate degrees (with most median earnings ranging from $20,000 to $50,000) and certificate degrees ($20,000 to $30,000). It should be further noted that there are 3 predominantly bachelor’s degree institutions whose median earnings 6 years after entry are substantially higher than those of other institutions at around $100,000 to $125,000. 

Furthermore, while the colleges in the data set - regardless of degree type - have relatively similar instructional expenditures, institutions with predominantly bachelor’s degrees do have greater variability in their instructional expenses per student compared to colleges with primarily associate or certificate programs. In particular, the instructional costs per student range from $307 to $30,243 in predominantly certificate degree colleges and from $621 to $61,772 in predominately associate degree institutions. Meanwhile, the instructional expenditures per full-time student for institutions that mostly award bachelor’s degrees range from $257 to $161,644, a much larger range compared to the other two college types. Nevertheless, for predominately bachelor’s degree institutions, high instructional costs per student are not necessarily associated with improved colleges’ median earnings. As seen from the plot, the three institutions with the highest median income after graduation have instructional expenses per student that are not very far from the mean, and many schools with very high instructional costs are in the same range of median income (around $50,000 to $75,000) as colleges with lower instructional expenditures per student.


## Fitted Model

I fit a multiple linear regression model to examine instructional costs affect salary after graduation. As a result, two variables, instructional expenses per full-time equivalent student and median earnings 6 years after entry, are included in the model. Average family income is also included as a precision variable: a person’s from a high-income family might gain better chances to work in a high-income job due to the cumulative effects of living in a wealthy, low-crime neighborhood, better family connections, overall better education from elementary school to college, etc. Meanwhile, a person’s family income does not affect the instructional expenditures of a college, so it functions as a precision variable. In addition, I center the instructional costs per student and the family income variable to make the intercept of the model more sensible to interpret because it does not make sense for a college to have no instructional expenses or for a family to have no income. 

The undergraduate degree the institution mainly awards is added to the model as a confounding variable. I use this variable as a confounder rather than an effect modifier because I believe the relationship between instructional expenses and median earnings after graduation does not differ depending on the type of degree a college predominantly awards. Nevertheless, the type of degree awarded affects median income, since the type of degree limits the types and levels of jobs a person is qualified for. Moreover, the type of degree awarded impacts instructional costs due to the length of time needed to finish different degrees, required facilities (e.g. research labs), etc. Considering that instructional expenditures do not affect the type of degree a college mainly awards, it is reasonable for a college’s predominantly degree type to act as a confounding variable. In the model, colleges that mainly award certificate degrees serve as the baseline metrics, whereas associate degree programs and bachelor’s degree institutions are displayed. 


```{r}
mod1 <- lm(MD_EARN_WNE_P6 ~ centeredexp + PREDDEG.Cat + centeredfaminc, data = scorecard) # fit a multiple linear regression model 

coef(mod1) %>% 
    format(scientific = FALSE) 

#confidence intervals
confint(mod1) %>% 
  format(scientific = FALSE)

#p-values
tidy(mod1)
```

The model demonstrates a positive relationship between instructional expenditures per student and institutional median earnings 6 years after entry. It also shows that people from families with high income are associated with greater chances to work in higher-income jobs after college. In addition, the median earnings for people with bachelor’s degrees are expected to be higher than that of people with associate and certificate degrees.

We estimate that the average median earnings 6 years after entry among students who go to colleges that primarily award certificate degrees and have an average instructional expense per student, and whose families earn an average income is $29,024. For institutions with the same predominant type of degree awarded and the same average family income, each dollar increase in instructional expenditures per student is associated with a $0.32 increase, on average, in median income among colleges. 

We are 95% confident that the true change in the average median income of colleges associated with each dollar increase in instructional expenses per student is between 0.29 and 0.36 dollars, holding average family income and the predominant type of degree the college awarded constant. In this context, the word “confident” refers to our interval construction process - the expectation that 95% of the samples will generate confidence intervals that contain the true population value. Our confidence interval is very small, suggesting that our sample is relatively large and has a lower degree of variability. In addition, the interval is comprised of only positive values, signaling that there is an associated increase in colleges’ median income post-graduation with greater instructional expenses, keeping average family earnings and the type of degree the college primarily awards constant. 

The p-value for the slope coefficient of the centered instructional expenses variable is smaller than 0.0001. This indicates that the probability of observing a test statistic as or more extreme than our observed test statistic for this variable is extremely small, assuming that the null hypothesis is true. Because our p-value is much lower than our chosen significance threshold of 0.05, we reject our null hypothesis that there is no difference in college’s median earnings among graduates 6 years after entry that is associated with different instructional expenditures per student, keeping average family income and the type of degree colleges predominantly award constant. Rather, we favor the alternative hypothesis that higher instructional expenses per student are associated with increased colleges’ median earnings among graduates 6 years after entry, accounting for average family income and the primary type of degree colleges award. 

Among schools with the same instructional fees per student and in families with similar income, students’ estimated median earnings in primarily associate-degree and bachelor-degree colleges are respectively, on average, $1,046 and $3,480 higher than that of students from predominantly certificate-degree colleges. Meanwhile, within institutions of similar type and similar instructional expenses per student, each dollar increase in the average family income of institutions is associated with a $0.17 rise, on average, in college median earnings. All of these variables have smaller p-values than our significant threshold, so we would favor the alternative hypotheses that there is a relationship between our explanatory variables and colleges’ median earnings, accounting for the other variables in the model.


## Model Evaluation

```{r, message=F}
#residuals vs fitted values
augment(mod1) %>%
    ggplot(aes(y = .resid, x = .fitted)) + 
    geom_point(alpha = 0.5) + 
    geom_smooth(se = FALSE, color = "#F699CD") + 
    geom_hline(yintercept = 0) + 
    labs(x = 'Fitted Values (USD)', y = 'Residuals (USD)', title = 'Residuals vs Fitted Values') + 
    theme_classic() +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

#residuals vs instructional expenses
augment(mod1) %>%
    ggplot(aes(y = .resid, x = centeredexp)) + 
    geom_point(alpha = 0.5) + 
    geom_smooth(se = FALSE, color = "#F699CD") + 
    geom_hline(yintercept = 0) + 
    labs(x = 'Centered Instructional Expenses (USD)', y = 'Residuals (USD)', title = 'Residuals vs Centered Instructional Expenses') +  
    theme_classic() +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

#residuals vs degree
augment(mod1) %>%
    ggplot(aes(y = .resid, x = PREDDEG.Cat)) + 
    geom_boxplot() + 
    geom_hline(yintercept = 0) + 
    labs(x = 'Predominant Degree Awarded', y = 'Residuals (USD)', title = 'Residuals vs Predominantly Awarded Degree Types') + 
    theme_classic() +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

#residuals vs average family income
augment(mod1) %>%
    ggplot(aes(y = .resid, x = centeredfaminc)) + 
    geom_point(alpha = 0.5) + 
    geom_smooth(se = FALSE, color = "#F699CD") + 
    geom_hline(yintercept = 0) + 
    scale_x_continuous(labels = scales :: comma) + 
    labs(x = 'Centered Average Family Income (USD)', y = 'Residuals (USD)', title = 'Residuals vs Centered Average Family Income') +  
    theme_classic() +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

#numerical summary
glance(mod1) 
```


For the plots with fitted values and centered average family income, the residual plots are quite straight, with many residuals close to 0 and have a relatively equal spread across prediction values, indicating that the model is quite consistent in its prediction. The residual boxplot with 3 types of predominantly awarded degrees also has a fairly equal spread, with the median values of the residuals across 3 categories very close to 0. Nevertheless, we can see that the residual range of colleges’ median earnings post-entry for predominantly certificate degree colleges is smaller than that of associate degree colleges, and that of associate degree colleges is smaller than primarily bachelor’s degree schools. This demonstrates that the model has a more consistent prediction for predominantly certificate degree colleges than associate and bachelor’s degree ones. Regarding the residual plot of the centered instructional expenses variable, the residuals are not mostly close to 0 but are more spread out, showing that the prediction, in this case, is not very consistent. 

The downside is that, according to the residual plot of fitted values, our model overpredicts for fitted values in the higher end. More specifically, when the fitted median earnings of colleges are above $50,000, our model overestimates by quite a large amount, at around $2,000 to $10,000, estimated from the graph. This is possibly due to the fact that most of the fitted values range from $20,000 to $50,000, whereas only a few of these values lie outside that range, leading to unreliable predictions. In addition, the boxplot with the primary degree awarded explanatory variables also indicates that the range of our value overestimation is wider than our value underestimation since the tail length for residual values greater than 0 is longer than that of residuals smaller than 0. In terms of outliers, the graphs with fitted values present that there are 3 outliers, more specifically the 3 fitted values with residuals greater than 60,000. As seen with the boxplots, these 3 outliers are overpredicted values when the primarily offered degree by colleges is bachelor’s degree. 

The R-squared of our model is 0.465, meaning that 46.5% of the variation in median earnings after entry can be explained by instructional expenses per full-time student, average family income, and the type of degree the colleges mainly award. Meanwhile, the model’s residual standard error is 7572, suggesting that our linear regression model can predict colleges’ median earnings after graduation with a margin of error of $15,150. This is quite a large amount of money, which might result from the impact of the outliers. Without the outliers, our model would be likely to have a smaller residual standard error and can generate more reliable predictions. Considering that our explanatory variables are able to account for almost half of the variance in colleges’ median earnings 6 years after entry, however, I think the model still can provide a relatively dependable prediction. 

# Part 4. Research Question 2 - Is a Private College Education Associated with a Higher Odds of Above-average Median Salary after College?

## Visualization

```{r, warning = FALSE}
scorecard %>%
    mutate(aboveaveincome = factor(aboveaveincome)) %>% 
    ggplot() +
    geom_mosaic(aes(x = product(aboveaveincome, Control.Cat), fill = aboveaveincome)) + 
    facet_grid( . ~ FAMINC.Cat) + 
    labs(title = "Odds of Institutions Having Above-Average Income", x = "Type of institution", y = "Have Above Average Income (1 = Yes, 0 = No)", fill = "Above Average Income (1 = Yes, 0 = No)") + 
    scale_fill_manual(values = c("#99C24D", "#048BA8")) + 
    theme_classic() +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

# Calculate percentage of public and private schools with above-average median income in each average family income group
scorecard %>%
    group_by(FAMINC.Cat, Control.Cat) %>%
    count(aboveaveincome) %>%
    mutate(ABI.perc = n*100 / sum(n))

# Calculate percentage of public and private schools within each average family income group
scorecard %>%
    group_by(FAMINC.Cat) %>%
    count(Control.Cat) %>%
    mutate(Control.perc = n*100 / sum(n))
```

It can be seen from the mosaic plot that the majority of colleges have a median income 6 years after entry lower than the average earnings in 2017. Private colleges have a greater proportion of schools with above-average median income compared to public schools across all three groups of colleges with different ranges of average family income per thousand. Specifically, for the group with the lowest average family income, the percentage of public schools with above-average median earnings is only 0.85%, whereas for private colleges, the percentage is 3.54%, about 4 times greater. With regard to institutions with average family income from 52.1 to 98.7 thousand USD, public schools with above-average median earnings comprise 4.77% of the total number of public institutions. Meanwhile, the percentage for private higher institutions is almost double that of public colleges at 8.78%. In terms of the highest average family income group, private colleges with above-average median earnings after graduation comprise one-fourth of the private school population in the data set, whereas for public institutions the number is 16.67%. These numbers also demonstrate that college institutions with higher average family income would also have a greater chance of having above-average median earnings among graduates: comparing the lowest and highest average family earnings group, the proportion of private schools with the above-average median income for the higher family income group is 7 times higher than that of schools having lower-income families. 

Furthermore, there is a higher proportion of private schools compared to public institutions within higher average family income groups. This difference can be observed most clearly in the group with average family earnings from 98.7 to 145 thousand dollars, in which the percentage of private schools is almost 90%. For institutions with average family income ranging from 52.1 to 98.7 thousand USD, private colleges still comprise the majority at 64.4%. However, regarding the colleges with the lowest average family income, it is the only group with a higher percentage of public schools than private institutions at 55.7%.

## Fitted Model

Because this research question involves a binary outcome, I fit a logistic regression model to restrict the predicted probabilities between 0 and 1. In order to decide what is considered above average income, I divide the median earning 6 years after entry variable into two categories based on its comparison to the 2017 average income of college graduates, which is $49,785 (Korn Ferry, 2017). For the variable on the control type of the institution, public colleges function as the baseline, while the coefficient for private colleges is shown in the model. 

The predictor variable involving the average family income of colleges functions as a confounder in the model. I believe that this variable is a confounding variable rather than an effect modifier since family income does not alter the impact that different institutional types have on income after graduation. Nonetheless, students from higher-income families would have a higher chance of going to private colleges than those from low-income backgrounds since private colleges often have higher tuition fees than public colleges. Furthermore, as mentioned previously, those from high-income families have greater access to high-income jobs due to better family connections and K-12 education, etc. Because whether a college is public or private does not impact a student’s family earnings, average family income meets the requirements of a confounder. I also modify the variable so that the model demonstrates family income in the thousand. I believe that this would make the interpretation of the variable’s slope more sensible since the effect of a $1 difference in average family income is likely to be negligible. 


```{r, message = F}
mod2 <- scorecard %>%
  with(glm(aboveaveincome ~ Control.Cat + FAMINC1000, family = binomial)) # fit a logistic regression model 

mod2 %>%
  coef() %>% 
  exp() 

#confidence intervals
confint(mod2) %>% 
  exp()

#p-values
tidy(mod2)
```

The model indicates that private colleges are associated with a greater odds of having an above-average median income among graduates compared to public universities, provided that the colleges being compared have a similar average family income. In particular, the odds of having above-average median earnings for public schools with an average family income of $0 are estimated to be only 0.004. Meanwhile, such odds are predicted to be 2.4 times higher for private colleges compared to the odds for public institutions, holding average family income constant. 

Based on our 95% confidence interval, our observed data would not be unusual if the true odds of having an above-average median income among private colleges is between 1.6 and 3.7 times higher compared to the odds for public colleges. This interval is quite narrow, suggesting that our sample is relatively large and has small variability. All the values contained in the interval are also greater than 1, suggesting that private colleges have a higher odds of having above-average median income among graduates compared to public universities, adjusting for average family income. When comparing two colleges of the same type which differ in average family income by $1,000, it is predicted that the ratio of odds of having above-average median earnings is 1.03, in which the group with higher average family income has the greater odds. 

The null hypothesis for our primary variable of interest is that the ratio of odds of having above-average median income between private and public universities is 1, holding average family income constant. From our data, the p-value for the exponentiated slope coefficient of the college’s control type variable is 3.48 × 10^(-5). This suggests that the probability of observing a test statistic as large or larger than our observed test statistic, assuming the null hypothesis is true, for our institutional control type variable is extremely small. Since our p-value is smaller than our chosen significance threshold of 0.05, we reject our null hypothesis that there is no relationship, keeping average family income constant, between the control type of higher education institutions and the odds of colleges having above-average median earnings among graduates. The p-value for our average family income variable is also smaller than 0.0001. As a result, our model is in favor of the alternative hypothesis that higher average family earnings are associated with an increased odds of colleges having above-average median income among graduates. 

## Model Evaluation

```{r}
threshold <- 0.05 

# predicted probability boxplots
mod2 %>%
  augment(type.predict = 'response') %>% 
  ggplot(aes(y = .fitted, x = factor(aboveaveincome))) + 
  geom_boxplot() + 
  ylab('Predicted Probability of Above Average Income') + 
  xlab('Actual Outcome of Colleges With Above Average Income (1 = Yes, 0 = No)') + 
  theme_classic() + 
  ggtitle('Predictions for Model 2') + 
  geom_hline(yintercept = threshold, color = 'red', linetype = 2) +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

# get binary predictions
mod2 %>%
  augment(type.predict = 'response') %>% 
  mutate(predictIncome = .fitted >= threshold) %>% 
  count(aboveaveincome, predictIncome) 
```

The boxplots are relatively separated: within the interquartile range, the predicted probabilities of colleges having above-average median income are mostly higher for those who were actually having above-average median earnings than for those who were not. To evaluate the model, I choose a threshold of 0.05 to balance the false negative and false positive rate of the model: estimating from the boxplot, this threshold gives correct prediction for approximately 75% of the cases on each side of the outcome. Based on my calculation, the accuracy percentage of the model is 73.2%. This indicates that the accuracy of the model is relatively high. The sensitivity rate is 68.5%, meaning that the false positive rate is 31.5%. In other words, among the institutions that have an above-average median income, our model is able to predict 68.5% of those cases correctly. Meanwhile, the specificity rate is 73.4%, so our false negative rate is 26.6%. To put it more simply, among the colleges with lower than average median income among graduates, our model can accurately predict 73.4% of those universities. 
